### 2021-07-29 doc/src/CSRs.tex
Replace CSRs *claimei with *topei
It was decided that standard RISC-V CSRs will not have side effects
on reads, so CSRs mclaimei and sclaimei are removed and replaced by
mtopei, stopei, and vstopei.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{4}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0xFB0} & MRO & \z{mtopi}    & Machine top interrupt \\
\hline
\multicolumn{4}{|c|}{Machine-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x358} & MRW & \z{mseteipnum}
                   & Machine set external interrupt-pending bit by number \\
\z{0x359} & MRW & \z{mclreipnum}
                   & Machine clear external interrupt-pending bit by number \\
\z{0x35A} & MRW & \z{mseteienum}
                   & Machine set external interrupt-enable bit by number \\
\z{0x35B} & MRW & \z{mclreienum}
                   & Machine clear external interrupt-enable bit by number \\
\z{0xFA8} & MRO & \z{mclaimei} & Machine claim top external interrupt \\
\hline
\multicolumn{4}{|c|}{Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x308} & MRW & \z{mvien}    & Machine virtual interrupt enables \\
\z{0x309} & MRW & \z{mvip}     & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{4}{|c|}{%
  Machine-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x313} & MRW & \z{midelegh} & Extension of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & \z{mieh}     & Extension of \z{mie} \\
\z{0x318} & MRW & \z{mvienh}   & Extension of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & \z{mviph}    & Extension of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & \z{miph}     & Extension of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{Machine-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-M}
\end{table*}

@@

Five CSRs, \z{mseteipnum}, \z{mclreipnum}, \z{mseteienum},
\z{mclreienum}, and \z{mclaimei}, interact with an IMSIC, so exist only
when an IMSIC is implemented.
Like the indirectly accessed IMSIC registers, these CSRs are documented
in Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0xDB0} & SRO & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x158} & SRW & \z{sseteipnum}
                 & Supervisor set external interrupt-pending bit by number \\
\z{0x159} & SRW & \z{sclreipnum}
                 & Supervisor clear external interrupt-pending bit by number \\
\z{0x15A} & SRW & \z{sseteienum}
                 & Supervisor set external interrupt-enable bit by number \\
\z{0x15B} & SRW & \z{sclreienum}
                 & Supervisor clear external interrupt-enable bit by number \\
\z{0xDA8} & SRO & \z{sclaimei} & Supervisor claim top external interrupt \\
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x114} & SRW & \z{sieh}     & Extension of \z{sie} \\
\z{0x154} & SRW & \z{siph}     & Extension of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{Supervisor-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-S}
\end{table*}

@@

Register \z{stopi} reports the highest-priority interrupt that
is pending and enabled for supervisor level, as specified in
Section~\ref{sec:stopi}.
The IMSIC CSRs, \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum},
\z{sclreienum}, and \z{sclaimei}, are described with the IMSIC in
Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name        & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Virtual Interrupts and Interrupt Priorities for VS Level} \\
\hline
\z{0x608} & HRW & \z{hvien}      & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & \z{hvictl}     & Hypervisor virtual interrupt control \\
\z{0x646} & HRW & \z{hviprio1}   & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & \z{hviprio2}   & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{4}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & \z{vsiselect}
                   & Virtual supervisor indirect register select \\
\z{0x251} & HRW & \z{vsireg}
                   & Virtual supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{VS-Level Interrupts} \\
\hline
\z{0xEB0} & HRO & \z{vstopi}     & Virtual supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{VS-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x258} & HRW & \z{vsseteipnum}
                   & VS set external interrupt-pending bit by number \\
\z{0x259} & HRW & \z{vsclreipnum}
                   & VS clear external interrupt-pending bit by number \\
\z{0x25A} & HRW & \z{vsseteienum}
                   & VS set external interrupt-enable bit by number \\
\z{0x25B} & HRW & \z{vsclreienum}
                   & VS clear external interrupt-enable bit by number \\
\hline
\multicolumn{4}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & \z{hidelegh}   & Extension of \z{hideleg} \\
\z{0x618} & HRW & \z{hvienh}     & Extension of \z{hvien} \\
\z{0x655} & HRW & \z{hviph}      & Extension of \z{hvip} \\
\z{0x656} & HRW & \z{hviprio1h}  & Extension of \z{hviprio1} \\
\z{0x657} & HRW & \z{hviprio2h}  & Extension of \z{hviprio2} \\
\z{0x214} & HRW & \z{vsieh}      & Extension of \z{vsie} \\
\z{0x254} & HRW & \z{vsiph}      & Extension of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{Hypervisor and VS CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-hypervisor}
\end{table*}

@@

Along the same lines, when \z{hstatus}.VGEIN is not the number of
an implemented guest external interrupt, attempts from \mbox{M-mode}
or \mbox{HS-mode} to access CSRs \z{vsseteipnum}, \z{vsclreipnum},
\z{vsseteienum}, or \z{vsclreienum} raise an illegal instruction
exception, and attempts from \mbox{VS-mode} to access \z{sseteipnum},
\z{sclreipnum}, \z{sseteienum}, \z{sclreienum}, or \z{sclaimei} raise a
virtual instruction exception.

>>>>

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{4}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0xFB0} & MRO & \z{mtopi}    & Machine top interrupt \\
\hline
\multicolumn{4}{|c|}{Machine-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x358} & MRW & \z{mseteipnum}
                   & Machine set external interrupt-pending bit by number \\
\z{0x359} & MRW & \z{mclreipnum}
                   & Machine clear external interrupt-pending bit by number \\
\z{0x35A} & MRW & \z{mseteienum}
                   & Machine set external interrupt-enable bit by number \\
\z{0x35B} & MRW & \z{mclreienum}
                   & Machine clear external interrupt-enable bit by number \\
\z{0x35D} & MRW & \z{mtopei}   & Machine top external interrupt \\
\hline
\multicolumn{4}{|c|}{Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x308} & MRW & \z{mvien}    & Machine virtual interrupt enables \\
\z{0x309} & MRW & \z{mvip}     & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{4}{|c|}{%
  Machine-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x313} & MRW & \z{midelegh} & Extension of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & \z{mieh}     & Extension of \z{mie} \\
\z{0x318} & MRW & \z{mvienh}   & Extension of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & \z{mviph}    & Extension of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & \z{miph}     & Extension of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{Machine-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-M}
\end{table*}

@@

Five CSRs, \z{mseteipnum}, \z{mclreipnum}, \z{mseteienum},
\z{mclreienum}, and \z{mtopei}, interact with an IMSIC, so exist only
when an IMSIC is implemented.
Like the indirectly accessed IMSIC registers, these CSRs are documented
in Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0xDB0} & SRO & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x158} & SRW & \z{sseteipnum}
                 & Supervisor set external interrupt-pending bit by number \\
\z{0x159} & SRW & \z{sclreipnum}
                 & Supervisor clear external interrupt-pending bit by number \\
\z{0x15A} & SRW & \z{sseteienum}
                 & Supervisor set external interrupt-enable bit by number \\
\z{0x15B} & SRW & \z{sclreienum}
                 & Supervisor clear external interrupt-enable bit by number \\
\z{0x15D} & SRW & \z{stopei}   & Supervisor top external interrupt \\
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x114} & SRW & \z{sieh}     & Extension of \z{sie} \\
\z{0x154} & SRW & \z{siph}     & Extension of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{Supervisor-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-S}
\end{table*}

@@

Register \z{stopi} reports the highest-priority interrupt that
is pending and enabled for supervisor level, as specified in
Section~\ref{sec:stopi}.
The IMSIC CSRs, \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum},
\z{sclreienum}, and \z{stopei}, are described with the IMSIC in
Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name        & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Virtual Interrupts and Interrupt Priorities for VS Level} \\
\hline
\z{0x608} & HRW & \z{hvien}      & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & \z{hvictl}     & Hypervisor virtual interrupt control \\
\z{0x646} & HRW & \z{hviprio1}   & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & \z{hviprio2}   & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{4}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & \z{vsiselect}
                   & Virtual supervisor indirect register select \\
\z{0x251} & HRW & \z{vsireg}
                   & Virtual supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{VS-Level Interrupts} \\
\hline
\z{0xEB0} & HRO & \z{vstopi}     & Virtual supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{VS-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x258} & HRW & \z{vsseteipnum}
                   & VS set external interrupt-pending bit by number \\
\z{0x259} & HRW & \z{vsclreipnum}
                   & VS clear external interrupt-pending bit by number \\
\z{0x25A} & HRW & \z{vsseteienum}
                   & VS set external interrupt-enable bit by number \\
\z{0x25B} & HRW & \z{vsclreienum}
                   & VS clear external interrupt-enable bit by number \\
\z{0x25D} & SRW & \z{vstopei}    & Virtual supervisor top external interrupt \\
\hline
\multicolumn{4}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & \z{hidelegh}   & Extension of \z{hideleg} \\
\z{0x618} & HRW & \z{hvienh}     & Extension of \z{hvien} \\
\z{0x655} & HRW & \z{hviph}      & Extension of \z{hvip} \\
\z{0x656} & HRW & \z{hviprio1h}  & Extension of \z{hviprio1} \\
\z{0x657} & HRW & \z{hviprio2h}  & Extension of \z{hviprio2} \\
\z{0x214} & HRW & \z{vsieh}      & Extension of \z{vsie} \\
\z{0x254} & HRW & \z{vsiph}      & Extension of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{Hypervisor and VS CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-hypervisor}
\end{table*}

@@

Along the same lines, when \z{hstatus}.VGEIN is not the number of
an implemented guest external interrupt, attempts from \mbox{M-mode}
or \mbox{HS-mode} to access CSRs \z{vsseteipnum}, \z{vsclreipnum},
\z{vsseteienum}, or \z{vsclreienum} raise an illegal instruction
exception, and attempts from \mbox{VS-mode} to access \z{sseteipnum},
\z{sclreipnum}, \z{sseteienum}, \z{sclreienum}, or \z{stopei} raise a
virtual instruction exception.

### 2021-07-29 doc/src/IMSIC.tex
Replace CSRs *claimei with *topei
It was decided that standard RISC-V CSRs will not have side effects
on reads, so CSRs mclaimei and sclaimei are removed and replaced by
mtopei, stopei, and vstopei.

For machine level, the relevant CSRs are:\nopagebreak
\begin{displayLinesTable}[l@{\qquad}l@{\qquad}l]
\z{miselect} & \z{mseteipnum} & \z{mclaimei} \\
\z{mireg}    & \z{mclreipnum} \\
             & \z{mseteienum} \\
             & \z{mclreienum} \\
\end{displayLinesTable}

When supervisor mode is implemented, the set of supervisor-level CSRs
matches those of machine level:
\begin{displayLinesTable}[l@{\qquad}l@{\qquad}l]
\z{siselect} & \z{sseteipnum} & \z{sclaimei} \\
\z{sireg}    & \z{sclreipnum} \\
             & \z{sseteienum} \\
             & \z{sclreienum} \\
\end{displayLinesTable}

And when the hypervisor extension is implemented, there is a
corresponding set of VS CSRs for all registers except \z{sclaimei}:
\begin{displayLinesTable}[l@{\qquad}l]
\z{vsiselect} & \z{vsseteipnum} \\
\z{vsireg}    & \z{vsclreipnum} \\
              & \z{vsseteienum} \\
              & \z{vsclreienum} \\
\end{displayLinesTable}

@@

Interrupt files at each level act identically.
For a given privilege level, values of the \z{*iselect} CSR in the
range \z{0x70}--\z{0xFF} select these registers of the corresponding
interrupt file:
\begin{displayLinesTable}[c@{\quad}l]
\z{0x70} & \z{eidelivery} \\
\z{0x72} & \z{eithreshold} \\
\z{0x76} & \z{topei} \\
\z{0x80} & \z{eip0} \\
\z{0x81} & \z{eip1} \\
\dots    & \ \dots \\
\z{0xBF} & \z{eip63} \\
\z{0xC0} & \z{eie0} \\
\z{0xC1} & \z{eie1} \\
\dots    & \ \dots \\
\z{0xFF} & \z{eie63} \\
\end{displayLinesTable}

Register numbers \z{0x71}, \z{0x73}--\z{0x75}, and \z{0x77}--\z{0x7F}
are reserved.
When a \z{*iselect} CSR has one of these values, reads from the
matching \emph{\texttt{*ireg} CSR} (\z{mireg}, \z{sireg}, or
\z{vsireg}) return zero, and writes to the \z{*ireg} CSR are ignored.
(For \z{vsiselect} and \z{vsireg}, all accesses depend on
\z{hstatus}.VGEIN being the valid number of a guest interrupt file.)

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Top external interrupt register (\zSafe{topei})}

\z{topei} is a read-only register whose value indicates the current
highest-priority pending-and-enabled interrupt that also exceeds the
priority threshold specified by \z{eithreshold}, if not zero.
Interrupts with lower identity numbers have higher priorities.

A read of \z{topei} returns zero either if no interrupt is both
pending in the \z{eip} array and enabled in the \z{eie} array, or if
\z{eithreshold} is not zero and no pending-and-enabled interrupt has an
identity number less than the value of \z{eithreshold}.
Otherwise, the value returned from a read of \z{topei} has this format:
\begin{displayLinesTable}[l@{\quad}l]
bits 26:16 & Interrupt identity \\
bits 10:0  & Interrupt priority (same as identity) \\
\end{displayLinesTable}
All other bit positions are zeros.

The interrupt identity reported in \z{topei} is the minor identity for
an external interrupt at the attached hart.

\begin{commentary}
The redundancy in the value read from \z{topei} is consistent with the
Advanced PLIC, which returns both an interrupt identity number and its
priority in the same format as above, but with the two components being
independent of one another.
\end{commentary}

Writes to \z{topei} are ignored.

@@

Several new machine-level CSRs interact directly with an IMSIC's
machine-level interrupt file:  \z{mseteipnum}, \z{mclreipnum},
\z{mseteienum}, \z{mclreienum}, and \z{mclaimei}.
If supervisor mode is implemented, there are several matching
supervisor CSRs that likewise interact with the supervisor-level
interrupt file:  \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum},
\z{sclreienum}, and \z{sclaimei}.
And if the hypervisor extension is implemented and field VGEIN of
\z{hstatus} is the number of an implemented guest interrupt file,
several VS CSRs interact with the chosen guest interrupt file:
\z{vsseteipnum}, \z{vsclreipnum}, \z{vsseteienum}, and \z{vsclreienum}.
These CSRs are documented by function below:

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Claim top external interrupt (\zSafe{*claimei})}

The \emph{\texttt{*claimei} CSRs}, \z{mclaimei} and \z{sclaimei}, are
read-only.

A read from \z{mclaimei} or \z{sclaimei} returns the same value as the
\z{topei} register of the respective interrupt file.
When this value is not zero, a read from \z{mclaimei} or \z{sclaimei}
has the simultaneous side effect of clearing the pending bit for the
reported interrupt identity in the interrupt file.

Although there is no VS CSR \z{vsclaimei}, if field VGEIN of
\z{hstatus} is the number of an implemented guest interrupt file,
reading \z{sclaimei} in \mbox{VS-mode} of course returns the same value
as the \z{topei} register of the selected guest interrupt file and,
if this value is not zero, clears the pending bit for the reported
interrupt identity in that interrupt file.

@@

When interrupt delivery is disabled by an interrupt file's
\z{eidelivery} register (\z{eidelivery} =~0), the interrupt signal from
the interrupt file is held de-asserted (false).
When interrupt delivery from the interrupt file is enabled
(\z{eidelivery} =~1), the interrupt signal from the interrupt file
is asserted if and only if reading register \z{topei} would return a
nonzero value.

A trap handler solely for external interrupts via an IMSIC could be
written roughly as follows:
\begin{displayLinesTable}
save processor registers \\
\z{i = }read CSR \z{mclaimei} or \z{sclaimei} \\
\z{i = i>>16} \\
call the interrupt handler for external interrupt \z{i} (minor identity) \\
restore processor registers \\
return from trap \\
\end{displayLinesTable}

>>>>

For machine level, the relevant CSRs are:\nopagebreak
\begin{displayLinesTable}[l@{\qquad}l@{\qquad}l]
\z{miselect} & \z{mseteipnum} & \z{mtopei} \\
\z{mireg}    & \z{mclreipnum} \\
             & \z{mseteienum} \\
             & \z{mclreienum} \\
\end{displayLinesTable}

When supervisor mode is implemented, the set of supervisor-level CSRs
matches those of machine level:
\begin{displayLinesTable}[l@{\qquad}l@{\qquad}l]
\z{siselect} & \z{sseteipnum} & \z{stopei} \\
\z{sireg}    & \z{sclreipnum} \\
             & \z{sseteienum} \\
             & \z{sclreienum} \\
\end{displayLinesTable}

And when the hypervisor extension is implemented, there is a
corresponding set of VS CSRs:
\begin{displayLinesTable}[l@{\qquad}l@{\qquad}l]
\z{vsiselect} & \z{vsseteipnum} & \z{vstopei} \\
\z{vsireg}    & \z{vsclreipnum} \\
              & \z{vsseteienum} \\
              & \z{vsclreienum} \\
\end{displayLinesTable}

@@

Interrupt files at each level act identically.
For a given privilege level, values of the \z{*iselect} CSR in the
range \z{0x70}--\z{0xFF} select these registers of the corresponding
interrupt file:
\begin{displayLinesTable}[c@{\quad}l]
\z{0x70} & \z{eidelivery} \\
\z{0x72} & \z{eithreshold} \\
\z{0x80} & \z{eip0} \\
\z{0x81} & \z{eip1} \\
\dots    & \ \dots \\
\z{0xBF} & \z{eip63} \\
\z{0xC0} & \z{eie0} \\
\z{0xC1} & \z{eie1} \\
\dots    & \ \dots \\
\z{0xFF} & \z{eie63} \\
\end{displayLinesTable}

Register numbers \z{0x71} and \z{0x73}--\z{0x7F}
are reserved.
When a \z{*iselect} CSR has one of these values, reads from the
matching \emph{\texttt{*ireg} CSR} (\z{mireg}, \z{sireg}, or
\z{vsireg}) return zero, and writes to the \z{*ireg} CSR are ignored.
(For \z{vsiselect} and \z{vsireg}, all accesses depend on
\z{hstatus}.VGEIN being the valid number of a guest interrupt file.)

@@

@@

Several new machine-level CSRs interact directly with an IMSIC's
machine-level interrupt file:  \z{mseteipnum}, \z{mclreipnum},
\z{mseteienum}, \z{mclreienum}, and \z{mtopei}.
If supervisor mode is implemented, there are several matching
supervisor CSRs that likewise interact with the supervisor-level
interrupt file:  \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum},
\z{sclreienum}, and \z{stopei}.
And if the hypervisor extension is implemented and field VGEIN of
\z{hstatus} is the number of an implemented guest interrupt file,
several VS CSRs interact with the chosen guest interrupt file:
\z{vsseteipnum}, \z{vsclreipnum}, \z{vsseteienum}, \z{vsclreienum},
and \z{vstopei}.
These CSRs are documented by function below:

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Top external interrupt (\zSafe{*topei})}

The value of a \emph{\texttt{*topei} CSR} (\z{mtopei}, \z{stopei}, or
\z{vstopei}) indicates the interrupt file's current highest-priority
pending-and-enabled interrupt that also exceeds the priority threshold
specified by its \z{eithreshold} register if \z{eithreshold} is not
zero.
Interrupts with lower identity numbers have higher priorities.

A read of a \z{*topei} CSR returns zero either if no interrupt
is both pending in the interrupt file's \z{eip} array and enabled
in its \z{eie} array, or if \z{eithreshold} is not zero and no
pending-and-enabled interrupt has an identity number less than the
value of \z{eithreshold}.
Otherwise, the value returned from a read of \z{*topei} has this
format:
\begin{displayLinesTable}[l@{\quad}l]
bits 26:16 & Interrupt identity \\
bits 10:0  & Interrupt priority (same as identity) \\
\end{displayLinesTable}
All other bit positions are zeros.

The interrupt identity reported in a \z{*topei} CSR is the minor
identity for an external interrupt at the hart.

\begin{commentary}
The redundancy in the value read from a \z{*topei} CSR is consistent
with the Advanced PLIC, which returns both an interrupt identity
number and its priority in the same format as above, but with the two
components being independent of one another.
\end{commentary}

A write to a \z{*topei} CSR \emph{claims} the reported interrupt
identity by clearing its pending bit in the interrupt file.
The value written is ignored;
rather, the current readable value of the register determines which
interrupt-pending bit is cleared.
Specifically, when a \z{*topei} CSR is written, if the register value
has interrupt identity $i$ in bits 26:16, then the interrupt file's
pending bit for interrupt~$i$ is cleared.
When a \z{*topei} CSR's value is zero, a write to the register has no
effect.

If a read and write of a \z{*topei} CSR are done together by a single
CSR instruction (CSRRW, CSRRS, or CSRRC), the value returned by the
read indicates the pending bit that is cleared.

\begin{commentary}
It is almost always a mistake to write to a \z{*topei} CSR without a
simultaneous read to learn which interrupt was claimed.
Note especially, if a read of a \z{*topei} register and a subsequent
write to the register are done by two separate CSR instructions, then
a higher-priority interrupt may become newly pending-and-enabled in the
interrupt file between the two instructions, causing the write to clear
the pending bit of the new interrupt and not the one reported by the
read.
Once the pending bit of the new interrupt is cleared, the interrupt is
lost.

If it is necessary first to read a \z{*topei} CSR and then subsequently
claim the interrupt as a separate step, the claim can be safely done by
writing the interrupt identity to the corresponding \z{*clreipnum} CSR,
instead of writing to \z{*topei}.
\end{commentary}

\begin{commentary}
Because the effect of a combined read-and-write of a \z{*topei} CSR
can be accomplished equally by a read of \z{*topei} followed by a write
of the interrupt identity to the corresponding \z{*clreipnum}, the
\z{*topei} CSRs could have been made read-only without the special
behavior defined for writes.
However, condensing these operations into a single CSR access intead of
two helps to reduce interrupt latency when critical.
\end{commentary}

@@

When interrupt delivery is disabled by an interrupt file's
\z{eidelivery} register (\z{eidelivery} =~0), the interrupt signal from
the interrupt file is held de-asserted (false).
When interrupt delivery from an interrupt file is enabled
(\z{eidelivery} =~1), its interrupt signal is asserted if and only
if the interrupt file has a pending-and-enabled interrupt that also
exceeds the priority threshold specified by \z{eithreshold}, if not
zero.

A trap handler solely for external interrupts via an IMSIC could be
written roughly as follows:
\begin{displayLinesTable}
save processor registers \\
\z{i = }read CSR \z{mtopei} or \z{stopei}, and write simultaneously to
 claim the interrupt \\
\z{i = i>>16} \\
call the interrupt handler for external interrupt \z{i} (minor identity) \\
restore processor registers \\
return from trap \\
\end{displayLinesTable}

The combined read and write of \z{mtopei} or \z{stopei} in the second
step can be done by a single CSRRW machine instruction,
\begin{displayLinesTable}
\z{csrrw }\textit{rd}\z{,} \z{mtopei} or \z{stopei,} \z{x0} \\
\end{displayLinesTable}
where \textit{rd} is the destination register for value~\z{i}.

### 2021-07-29 doc/src/VSLevel.tex
Replace CSRs *claimei with *topei
It was decided that standard RISC-V CSRs will not have side effects
on reads, so CSRs mclaimei and sclaimei are removed and replaced by
mtopei, stopei, and vstopei.

When \z{hstatus}.VGEIN is the valid number of a guest interrupt
file, values of \z{vsiselect} in the range \z{0x70}--\z{0xFF} select
registers of this guest interrupt file, just as values of \z{siselect}
in the same range select registers of the IMSIC's true supervisor-level
interrupt file.
The registers of an interrupt file that are accessed indirectly through
\z{vsiselect} and \z{vireg} are documented in Chapter~\ref{ch:IMSIC}
on the IMSIC, along with the effects of accessing CSRs \z{vsseteipnum},
\z{vsclreipnum}, \z{vsseteienum}, and \z{vsclreienum}.
Because all IMSIC interrupt files act identically, the guest interrupt
file that a virtual hart accesses through CSRs \z{siselect}, \z{sireg},
\z{sclaimei}, \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum}, and
\z{sclreienum} is indistinguishable from a true supervisor-level
interrupt file as seen from \mbox{S-mode} (or \mbox{HS-mode}).

@@

\item
if \z{hstatus}.VGEIN is the valid number of a guest interrupt file,
bit~9 is one in both \z{vsip} and \z{vsie}, and the \z{topei} register
of the selected guest interrupt file is not zero:
\begin{tightList}
\item[+]
a supervisor external interrupt (code~9) with the priority number
indicated by the guest interrupt file's \z{topei} register;
\end{tightList}

\item
if \z{hstatus}.VGEIN is the valid number of a guest interrupt file,
bit~9 is one in both \z{vsip} and \z{vsie}, and the \z{topei} register
of the selected guest interrupt file is zero:
\begin{tightList}
\item[+]
a supervisor external interrupt (code~9) with priority number 256;
\end{tightList}

>>>>

When \z{hstatus}.VGEIN is the valid number of a guest interrupt
file, values of \z{vsiselect} in the range \z{0x70}--\z{0xFF} select
registers of this guest interrupt file, just as values of \z{siselect}
in the same range select registers of the IMSIC's true supervisor-level
interrupt file.
The registers of an interrupt file that are accessed indirectly through
\z{vsiselect} and \z{vireg} are documented in Chapter~\ref{ch:IMSIC}
on the IMSIC, along with the effects of accessing CSRs \z{vsseteipnum},
\z{vsclreipnum}, \z{vsseteienum}, and \z{vsclreienum}.
Because all IMSIC interrupt files act identically, the guest interrupt
file that a virtual hart accesses through CSRs \z{siselect}, \z{sireg},
\z{sseteipnum}, \z{sclreipnum}, \z{sseteienum}, \z{sclreienum},
and \z{stopei} is indistinguishable from a true supervisor-level
interrupt file as seen from \mbox{S-mode} (or \mbox{HS-mode}).

@@

\item
if \z{hstatus}.VGEIN is the valid number of a guest interrupt file,
bit~9 is one in both \z{vsip} and \z{vsie}, and \z{vstopei}
is not zero:
\begin{tightList}
\item[+]
a supervisor external interrupt (code~9) with the priority number
indicated by \z{vstopei};
\end{tightList}

\item
if \z{hstatus}.VGEIN is the valid number of a guest interrupt file,
bit~9 is one in both \z{vsip} and \z{vsie}, and \z{vstopei} is zero:
\begin{tightList}
\item[+]
a supervisor external interrupt (code~9) with priority number 256;
\end{tightList}
